\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}

\title{EE 5103 --- Final Programming Project\\
\large Circuit Analysis Library in C++}
\author{Jordan Cavlovic}
\date{May 2026}

\begin{document}

\maketitle

\section{Problem Description}

The goal of this project is to build a small C++ library that represents
simple AC circuits, computes their equivalent complex impedance, the
phasor current drawn from the source, the voltage drop across each
component, and the average real power dissipated. The user can construct
resistors, capacitors, inductors, and ideal voltage sources, group them
into series or parallel circuits, combine entire circuits with the
\texttt{+} and \texttt{*} operators, and save or load circuits as text
files. The DC case falls out for free: at $\omega = 0$ an inductor's
impedance collapses to zero and a capacitor's blows up to infinity, so
the same analysis covers both regimes without a separate code path.

Hand-computing impedances, phasor currents, and dissipated power becomes
tedious past a few elements, so a small object-oriented library is a
better fit: each element type knows how to compute its own impedance,
voltage drop, and power, and a container class composes them.

\section{Example Run}

The driver prompts the user for a source voltage $V_s$ (RMS), a source
frequency $f$ (Hz, with $0$ meaning DC), three resistor values, a
capacitor value $C_1$ (F), and an inductor value $L_1$ (H). It then
builds a series circuit and reports the total impedance magnitude and
angle, the phasor current, and the voltage drop and power for each
element. The same source is then applied to a parallel network, and the
program also reloads a flat circuit from \texttt{circuit.txt} to exercise
the file I/O path. The capacitor and inductor are included in both
circuits so the user can see the DC steady-state behavior fall out of
the general analysis at $f = 0$: the capacitor reads as an open and the
inductor as a perfect wire, so a series network with a capacitor in it
has $|Z| = \infty$ and zero current, while a parallel network with an
inductor in it shorts to $|Z| = 0$. The resistor-only tables below
isolate the meaningful drops and powers for the DC case; the storage
elements contribute zero voltage drop and zero dissipated power when
$\omega = 0$.

\begin{table}[h]
\centering
\caption{Series-circuit summary for three sample runs at $f = 0$ (DC), where
$|Z|$ is the sum of the three resistors.}
\begin{tabular}{@{}rrrrrrr@{}}
\toprule
$V_s$ & $R_1$& $R_2$& $R_3$
& $|Z|$& $|I|$& $P_{total}$\\
\midrule
12 & 100 & 220 & 470 & 790 & 0.01519 & 0.182 \\
24 & 100 & 200 & 300 & 600 & 0.04000 & 0.960 \\
 5 &  10 &  20 &  30 &  60 & 0.08333 & 0.417 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Per-element voltage drop magnitude and average real power for the
first run ($V_s = 12$~V, $f = 0$, series).}
\begin{tabular}{@{}lrrr@{}}
\toprule
Element & $R$& $|V_{drop}|$& $P$\\
\midrule
$R_1$ & 100 & 1.519 & 0.023 \\
$R_2$ & 220 & 3.342 & 0.051 \\
$R_3$ & 470 & 7.139 & 0.108 \\
\midrule
Total & 790 & 12.000 & 0.182 \\
\bottomrule
\end{tabular}
\end{table}

The next two tables exercise the same series network at $f = 60$~Hz so the
capacitor and inductor contribute non-zero reactance. The voltage drops
across $C_1$ and $L_1$ are no longer zero, and they carry the expected
$\pm 90^{\circ}$ phase relative to the current. Their average real power
remains zero because ideal storage elements only exchange energy with the
source.

\begin{table}[h]
\centering
\caption{AC series-circuit summary for two sample runs at $f = 60$~Hz.
$X_C = 1/(\omega C_1)$ and $X_L = \omega L_1$.}
\begin{tabular}{@{}rrrrrrrrr@{}}
\toprule
$V_s$ & $R_1$& $R_2$& $R_3$& $C_1$ (F)& $L_1$ (H)
& $|Z|$& $\angle Z$ (deg)& $|I|$\\
\midrule
 12 & 100 & 220 & 470 & $100\!\times\!10^{-6}$ & 0.05 & 790.037 & $-0.557$ & 0.01519 \\
120 & 100 & 200 & 300 & $ 10\!\times\!10^{-6}$ & 0.50 & 604.890 & $-7.291$ & 0.19838 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Per-element voltage drop magnitude/angle and average real power
for the second AC series run ($V_s = 120$~V, $f = 60$~Hz,
$C_1 = 10\,\mu\mathrm{F}$, $L_1 = 0.5$~H). The capacitor lags the current
by $90^{\circ}$ and the inductor leads it by $90^{\circ}$.}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Element & Value & $|V_{drop}|$ (V) & $\angle V_{drop}$ (deg) & $P$ (W)\\
\midrule
$R_1$ & 100~$\Omega$        & 19.838 &   7.291 &  3.936 \\
$R_2$ & 200~$\Omega$        & 39.677 &   7.291 &  7.871 \\
$R_3$ & 300~$\Omega$        & 59.515 &   7.291 & 11.807 \\
$C_1$ & 10~$\mu$F           & 52.623 & $-82.709$ &  0.000 \\
$L_1$ & 0.5~H               & 37.394 &  97.291 &  0.000 \\
\midrule
Total & ---                 & 120.000 & 0.000 & 23.613 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Per-branch current magnitude/angle and average real power for the
parallel network on the second AC run ($V_s = 120$~V, $f = 60$~Hz,
$C_1 = 10\,\mu\mathrm{F}$, $L_1 = 0.5$~H). Each branch sees the full
source voltage; the capacitor branch leads by $90^{\circ}$ and the
inductor branch lags by $90^{\circ}$.}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Element & Value & $|I_{branch}|$ (A) & $\angle I_{branch}$ (deg) & $P$ (W)\\
\midrule
$R_1$ & 100~$\Omega$ & 1.200 &   0.000 & 144.000 \\
$R_2$ & 200~$\Omega$ & 0.600 &   0.000 &  72.000 \\
$R_3$ & 300~$\Omega$ & 0.400 &   0.000 &  48.000 \\
$C_1$ & 10~$\mu$F    & 0.452 &  90.000 &   0.000 \\
$L_1$ & 0.5~H        & 0.637 & $-90.000$ &   0.000 \\
\midrule
Total & ---          & 2.208 & $-4.787$ & 264.000 \\
\bottomrule
\end{tabular}
\end{table}

\section{Software Design and Class Structure}

The design uses the composite pattern, which means a circuit can act like a single component. This lets us use one recursive type system to handle both individual elements and groups of elements without needing separate logic for each case.
At the core is an abstract base class called Component. It declares a few pure virtual methods — impedance(omega), type(), clone(), and print() — and stores a label and value that all components share. It also provides default implementations of voltageDrop(I, omega) and power(I, omega) using the phasor formulas $V = IZ$ and $P = |I|^2\,\mathrm{Re}(Z)$. Subclasses only need to override these when the actual electrical behavior is different from the basic passive-element model.
There are four classes: Resistor, Capacitor, Inductor, and VoltageSource. Each one reports its complex impedance differently — a resistor gives $R$, an inductor gives $j\omega L$, a capacitor gives $1/(j\omega C)$ (which becomes infinity at $\omega = 0$ so a DC analysis sees an open), and an ideal voltage source is treated as 0. VoltageSource overrides voltageDrop and power to match its real behavior: it reports a negative drop and negative dissipation to show it's supplying power rather than consuming it. The passive elements just inherit the default phasor implementations, so reactive components automatically come out with the right magnitude and angle.
Circuit also extends Component, so it can be nested inside other circuits seamlessly. It holds a list of child components and knows whether they're connected in series or parallel. From there it calculates the total impedance, summing for series, and using the reciprocal-of-admittance formula for parallel. It also has an analyze() method that prints out the total impedance magnitude and angle, the applied voltage and angular frequency, the phasor current, and the voltage drop magnitude/angle and real power for each component. When it hits a nested sub-circuit, it uses a dynamic cast to recurse into it with the right branch voltage.
Circuit also supports operator+ for series combinations and operator* for parallel combinations, along with save() and load() for writing and reading circuits from a text file. To keep the load function clean, a helper class called ComponentFactory handles mapping type-name strings to the right concrete objects, so Circuit doesn't need to know anything about the specific leaf classes.
Everything gets wired together in main.cpp. Since every node in the tree exposes the same interface, client code never has to worry about whether it's dealing with a single resistor or a deeply nested sub-circuit.

\section{Discussion of C++ Concepts Used}

\paragraph{Inheritance and runtime polymorphism.}
\texttt{Component} declares pure virtual methods that every concrete subclass
has to override. \texttt{Circuit} stores its children as
\texttt{std::unique\_ptr<Component>}, so a single virtual call figures out
the right implementation at runtime, whether the child is a plain
resistor or a nested sub-circuit. The base destructor is marked
\texttt{virtual} so that deleting a derived object through a base pointer
doesn't cause undefined behavior.

\paragraph{Composite pattern.}
Because \texttt{Circuit} itself inherits from \texttt{Component}, a
circuit can contain other circuits. The recursion basically handles itself:
\texttt{Circuit::impedance} just calls \texttt{impedance} on each child
and lets virtual dispatch do the work, and \texttt{Circuit::analyze} does
the same with \texttt{voltageDrop} and \texttt{power}.
\texttt{analyze} also uses \texttt{dynamic\_cast} to check whether a
child is itself a \texttt{Circuit}, so it can recurse with the right
branch voltage. This same idea is what makes \texttt{operator+} and
\texttt{operator*} work, they wrap clones of both operands inside a
new series or parallel circuit.

\paragraph{Smart pointers and RAII.}
The library doesn't use any raw owning pointers. Children are stored as
\texttt{std::unique\_ptr<Component>}, so memory is cleaned up automatically
when the owning vector goes out of scope, and there's no ambiguity about
who owns what. Objects are created through \texttt{std::make\_unique}. File
handles use \texttt{std::ifstream} / \texttt{std::ofstream}, so streams
close on their own even if something throws an exception.

\paragraph{Virtual cloning.}
Deep copies go through the virtual \texttt{clone()} method on each
subclass, which avoids object slicing when copying a polymorphic
container. \texttt{Circuit::clone} just recursively clones every child.

\paragraph{Operator overloading.}
\texttt{Circuit::operator+} and \texttt{Circuit::operator*} let the client
write \texttt{series + parallel} and \texttt{parallel * parallel}, which
mirrors the actual algebraic rules for circuits. \texttt{operator<<} is
overloaded as a friend free function on \texttt{Component} and delegates
to the virtual \texttt{print} method, so any node can be printed with the
same syntax.

\paragraph{Templates and the STL.}
The implementation relies on \texttt{std::vector}, \texttt{std::unique\_ptr},
\texttt{std::string}, \texttt{std::ifstream} / \texttt{std::ofstream}, and
\texttt{std::complex<double>} (aliased as \texttt{Complex}) to carry
phasor impedances, currents, and voltages. \texttt{std::numeric\_limits<double>::infinity()}
represents a DC-blocking capacitor, and \texttt{<cmath>}'s \texttt{std::isinf}
and \texttt{std::isfinite} are used to guard the series sum and parallel
reciprocal so that an open branch propagates correctly without producing
NaN. \texttt{std::abs}, \texttt{std::arg}, and \texttt{std::norm} on
\texttt{Complex} pull out the magnitude, angle, and squared magnitude
that the analysis prints.

\paragraph{Run-time type identification.}
\texttt{Circuit::analyze} uses \texttt{dynamic\_cast<const Circuit*>} on
each child pointer to decide whether to recurse into a sub-circuit with
its branch voltage; leaf components don't need the recursive call. This
keeps the analysis working correctly at any nesting depth without having
to add a virtual hook to the leaf interface.

\paragraph{Strongly-typed enums.}
\texttt{Circuit::Topology} is declared as \texttt{enum class}, so its
enumerators don't pollute the surrounding scope and can't be accidentally
converted to integers.

\paragraph{Move semantics.}
Constructors take string parameters by value and \texttt{std::move} them
into the member. \texttt{Circuit::add} accepts a \texttt{std::unique\_ptr}
by value and \texttt{std::move}s it into the vector, making the ownership
transfer obvious at the call site.

\paragraph{Factory pattern.}
\texttt{ComponentFactory::create} maps a type-name string to the right
concrete object, so \texttt{Circuit::load} doesn't need to know anything
about the leaf classes. Adding a new component type only requires touching
the factory and the new class itself.

\paragraph{Exceptions.}
\texttt{Circuit::save} and \texttt{Circuit::load} throw
\texttt{std::runtime\_error} when the file can't be opened, and the
factory throws \texttt{std::invalid\_argument} for an unknown type name,
giving the caller a clear signal that something went wrong.

\paragraph{Const correctness.}
Methods that don't modify state are marked \texttt{const}. Read-only
access to the child vector is exposed through a \texttt{const\&} getter,
which stops callers from accidentally mutating the internal list while
still allowing efficient iteration.

\section{Summary}

The project covers all of its functional goals of building, analyzing,
combining, saving, and loading AC circuits, and reporting per-component
voltage drop magnitude/angle and average real power dissipation, with the
DC case handled as the $\omega = 0$ specialization of the same analysis.
The composite pattern keeps the data model consistent across all levels,
smart pointers make ownership unambiguous, operator overloading gives the
user natural algebraic syntax, and the factory keeps the loader decoupled
from the concrete element types.

\end{document}
